---
title: "tauBayesW: Bayesian Weighted Quantile Regression"
author: "Tomas Rodriguez Taborda"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tauBayesW}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

# tauBayesW: An R Framework for Bayesian Weighted Quantile Regression

[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/tauBayesW)](https://cran.r-project.org/package=tauBayesW)
[![downloads](https://cranlogs.r-pkg.org/badges/tauBayesW)](https://cran.r-project.org/package=tauBayesW)

## What is tauBayesW?

**tauBayesW** is an R package that provides a comprehensive framework for Bayesian weighted quantile regression with survey designs. The package implements state-of-the-art MCMC and EM algorithms to handle complex survey data while maintaining computational efficiency and statistical rigor.

Unlike traditional quantile regression approaches, tauBayesW:
- 🎯 **Handles survey weights naturally** in the Bayesian framework
- 🔄 **Implements multiple MCMC methods** (ALD, Score, Approximate)  
- 📊 **Estimates multiple quantiles efficiently** using EM algorithms
- 🎨 **Provides comprehensive visualization** tools
- 📈 **Includes model diagnostics** and convergence checking

## Installation

```{r installation, eval=FALSE}
# From CRAN (when available)
install.packages("tauBayesW")

# From GitHub (development version)
# remotes::install_github("torodriguezt/tauBayesW")
```

```{r setup}
library(tauBayesW)
set.seed(123)
```

## Citation

To cite tauBayesW in publications, please use:

```{r citation, eval=FALSE}
citation("tauBayesW")
```

## Getting Started

tauBayesW provides three main functions that work together to cover all aspects of Bayesian quantile regression:

- **`bqr.svy()`**: 🎯 Bayesian quantile regression for single quantiles with survey weights
- **`mo.bqr.svy()`**: 📊 Multiple quantile regression using efficient EM algorithms  
- **`plot_quantile()`**: 🎨 Comprehensive visualization functions for quantile regression

### Basic Usage

Let's start with a simple example using survey data:

```{r basic-example}
# Simulate survey data with design weights
n <- 500
x1 <- rnorm(n)
x2 <- rbinom(n, 1, 0.4)
y <- 1 + 2*x1 + 1.5*x2 + rnorm(n)
weights <- runif(n, 0.5, 2)  # Survey weights

data <- data.frame(y, x1, x2)

# Single quantile regression (median)
model_median <- bqr.svy(y ~ x1 + x2, 
                        data = data,
                        weights = weights,
                        quantile = 0.5,
                        n_mcmc = 2000,
                        verbose = FALSE)

summary(model_median)
```

### Multiple Quantile Analysis

```{r multiple-quantiles}
# Multiple quantiles simultaneously
quantiles <- c(0.1, 0.25, 0.5, 0.75, 0.9)

model_multi <- mo.bqr.svy(y ~ x1 + x2,
                          data = data, 
                          weights = weights,
                          quantile = quantiles,
                          max_iter = 100,
                          verbose = FALSE)

summary(model_multi)
```

### Visualization

```{r visualization, fig.width=8, fig.height=6}
# Create univariate data for better visualization
n_vis <- 300
x_vis <- seq(-2, 2, length.out = n_vis)
y_vis <- 1 + 2*x_vis + (0.5 + 0.3*abs(x_vis))*rnorm(n_vis)  # Heteroscedastic
weights_vis <- rgamma(n_vis, 2, 1)
data_vis <- data.frame(y = y_vis, x = x_vis)

# Fit single quantile for plotting
model_vis <- bqr.svy(y ~ x, 
                     data = data_vis,
                     weights = weights_vis,
                     quantile = 0.5,
                     n_mcmc = 2000,
                     verbose = FALSE)

# Plot quantile regression with data points
plot_quantile_with_points(model_vis, data = data_vis, predictor = "x",
                          point_alpha = 0.6,
                          main = "Bayesian Quantile Regression (τ = 0.5)")
```

## Advanced Features

### 1. Multiple MCMC Methods

tauBayesW implements three different MCMC approaches:

```{r mcmc-methods}
n <- 800
x1 <- runif(n)
x2 <- rnorm(n)
y <- 3 + 2*x1 + 1.5*x2 + rnorm(n)
weights <- rep(1, n)
data <- data.frame(y, x1, x2)

methods <- c("ald", "score", "approximate")
results <- list()

for(method in methods) {
  results[[method]] <- bqr.svy(y ~ x1 + x2,
                               data = data,
                               weights = weights,
                               quantile = 0.3,
                               method = method,
                               n_mcmc = 60000,
                               burnin = 25000,
                               verbose = FALSE)
}

coef_names <- c("(Intercept)", "x1", "x2")
coef_comparison <- do.call(rbind, lapply(results, function(mod) {
  coefs <- coef(mod)
  as.numeric(coefs[coef_names])
}))
colnames(coef_comparison) <- coef_names
rownames(coef_comparison) <- methods
print(round(coef_comparison, 4))

```

### 2. Complex Survey Designs

```{r survey-design}
# Simulate stratified survey data
n_strata <- 3
strata_data <- do.call(rbind, lapply(1:n_strata, function(s) {
  n_s <- 150
  x1 <- rnorm(n_s, mean = s, sd = 1)
  x2 <- rbinom(n_s, 1, prob = 0.2 + 0.2*s)
  y <- 1.5 + 2*x1 + 1.2*x2 + rnorm(n_s, sd = 0.8 + 0.2*s)
  
  # Inverse probability weights
  weights <- 1 / (0.1 + 0.3*s/n_strata)
  
  data.frame(y, x1, x2, stratum = s, weights = weights)
}))

# Analyze stratified data
strata_model <- bqr.svy(y ~ x1 + x2,
                        data = strata_data,
                        weights = strata_data$weights,
                        quantile = 0.5,
                        n_mcmc = 2000,
                        verbose = FALSE)

summary(strata_model)
```


### 4. Multiple Quantile Visualization

```{r multi-quantile-viz, fig.width=8, fig.height=6}
# Fit multiple quantiles for univariate case
model_multi_vis <- mo.bqr.svy(y ~ x,
                              data = data_vis,
                              weights = weights_vis,
                              quantile = c(0.1, 0.25, 0.5, 0.75, 0.9),
                              max_iter = 100,
                              verbose = FALSE)

# Plot multiple quantile curves
plot(model_multi_vis, type = "quantiles")

# Plot coefficient evolution across quantiles  
plot(model_multi_vis, type = "convergence")

```





## Methodology

### Theoretical Background

The package implements Bayesian quantile regression using the **Asymmetric Laplace Distribution (ALD)** representation:

$$y_i \sim ALD(\mu_i, \sigma, \tau)$$

where $\mu_i = \mathbf{x}_i^T \boldsymbol{\beta}$ and the likelihood is:

$$f(y_i|\mu_i,\sigma,\tau) = \frac{\tau(1-\tau)}{\sigma} \exp\left(-\rho_\tau\left(\frac{y_i-\mu_i}{\sigma}\right)\right)$$

The check function $\rho_\tau(u) = u(\tau - \mathbb{I}(u < 0))$ naturally incorporates quantile-specific loss.

### Survey Weights Integration

Survey weights $w_i$ are naturally incorporated in the Bayesian framework through the weighted likelihood:

$$L(\boldsymbol{\beta}, \sigma) = \prod_{i=1}^n f(y_i|\mu_i, \sigma, \tau)^{w_i}$$

This approach maintains design consistency while providing full posterior inference.

### EM Algorithm for Multiple Quantiles

For computational efficiency with multiple quantiles, the package implements an EM algorithm that:

1. **E-step**: Computes posterior expectations of latent variables
2. **M-step**: Updates parameters via weighted optimization

This ensures:
- ✅ Computational efficiency
- ✅ Numerical stability  

## Simulation and Validation

```{r simulation}
# Built-in simulation function
sim_data <- simulate_mo_bqr_data(n = 700,
                                 beta = c(1, 2, -0.5),
                                 seed = 456)

# Fit model to simulated data
sim_model <- mo.bqr.svy(y ~ x1 + x2,
                        data = sim_data$data,
                        weights = sim_data$weights,
                        quantile = c(0.25, 0.5, 0.75),
                        verbose = FALSE)

# Compare true vs estimated coefficients
true_coefs <- c(1, 2, -0.5)
estimated_coefs <- coef(sim_model)

print("True coefficients:")
print(true_coefs)
print("Estimated coefficients quantile 0.25:")
print(estimated_coefs[1]) 
```

## Additional Documentation

For comprehensive documentation, examples, and advanced tutorials, visit:

**📖 [https://torodriguezt.github.io/tauBayesW/](https://torodriguezt.github.io/tauBayesW/)**

## Dependencies

tauBayesW is designed to be lightweight with minimal dependencies:

- **Base R packages**: stats, grDevices, graphics
- **C++ integration**: Rcpp, RcppEigen, RcppArmadillo
- **Suggested**: survey (for complex survey examples)

This makes tauBayesW safe to use as a dependency in other packages without dependency conflicts.

## Package Functions Overview

| Function | Purpose | Type |
|----------|---------|------|
| `bqr.svy()` | Single quantile regression | Main |
| `mo.bqr.svy()` | Multiple quantile regression | Main |
| `plot_quantile()` | Quantile curve visualization | Plotting |
| `plot_quantile_with_points()` | Scatter plot with quantile overlay | Plotting |
| `convergence_check()` | MCMC convergence diagnostics | Diagnostic |
| `simulate_mo_bqr_data()` | Data simulation for testing | Utility |

All functions come with comprehensive S3 methods for `print()`, `summary()`, `plot()`, `coef()`, `fitted()`, `residuals()`, and `logLik()`.

## References
   - Yu, K. and Moyeed, R. A. (2001). Bayesian quantile regression. *Statistics & Probability Letters*, 54(4), 437-447.
   - Kozumi, H. and Kobayashi, G. (2011). Gibbs sampling methods for Bayesian quantile regression. *Journal of Statistical Computation and Simulation*, 81(11), 1565-1578.
   - Lohr, S. L. (2009). *Sampling: Design and Analysis*. Chapman and Hall/CRC.
   - Gelman, A., et al. (2013). *Bayesian Data Analysis*. Chapman and Hall/CRC.

---
